A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-04-18 masters_thesis/si_paper/paper.tex
Updated the journal paper version of my thesis

We ran experiments to compare our speculative parallelization of PSO to the
standard parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The exact evaluation time at which this
architecture becomes reasonable depends on the amount of communication overhead
in the parallel implementation and the number of particles in the swarm.  For
our implementation we found that time to be around 100 milliseconds for swarms
of 50 particles.  When the swarm size increases, the minimum evaluation time at
which this architecture should be used decreases.

@@

Just as serial PSO papers do not report actual running times of their specific
implementations, we do not report running times, favoring the abstraction of
time steps.  That being said, given the time required for the evaluation of the
objective function and the communication overhead per iteration for a specific
implementation, an actual running time can be recovered from all of our
results.  Simply multiply the number of time steps by the sum of the function
evaluation time and the overhead.  We wish to stress that the ``time steps'' we
report are proportional to wall-clock time, given the architecture we have
assumed.

@@

One might question the use of benchmark functions for evaluating our methods,
as they take on the order of 1 millisecond to evaluate instead of the 100s that
we said were required to justify our parallel architecture.  The purpose of
exploring the behavior of an algorithm on benchmark functions, whether in
serial or in parallel optimization research, is because benchmark functions are
believed to represent functions that one might encounter in the real world,
regardless of the amount of time those functions take to evaluate.  While
benchmark functions themselves take fractions of a second to evaluate and thus
have no need of the kind of parallelization we explore in this paper, they also
have no need of particle swarm optimization at all, as they can be solved
analytically.  Yet they are useful to us and to optimization researchers
generally because they stand as surrogates for the kinds of functions
practitioners are actually interested in, and they allow us to explore the
behavior of optimization algorithms on several different classes of functions
in an easy and standardized way.

To aid in convincing the reader that performance on benchmark functions roughly
corresponds to performance on real-world problems, we also tested our
parallelization methods on a typical research problem, that of fitting a model
to a large quantity of data.  We generated 10,000 data points from a radial
basis function network with 10 bases and some added Gaussian noise.  We then
fit a radial basis function network to the data using PSO.  This amounted to a
30 dimensional function to optimize, with a function evaluation time on the
order of two seconds.  We refer to this function as ``the model fitting
problem.''

@@

\subsubsection{Sphere}

@@

Note here that \fig{griewank-manyiters} is interesting to compare to
\fig{basic-griewank1}.  The ordering of the methods in terms of the number of
time steps to completion is the same in both figures.  What is different is
that at 240 processors, most of our methods fail to find the optimum 100\% of
the time, while at 800 processors, all but Many Iterations succeed 100\% of the
time, and Many Iterations is very close.  At some point between 240 and 800
processors, SEPSO and Pick Best become successful, and at that point it is by
far better to use speculative methods than standard parallelizations.  We can
also assume that at some point above 800 processors, Many Iterations will be
become successful and will be the preferred method.  Once there are enough
particles in the swarm to guarantee success when optimizing a function, the
best use of additional processors is to speculate as far ahead as possible,
speeding up the progress of the algorithm.

@@

>>>>

We ran experiments to compare our speculative parallelization of PSO to the
standard parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation.  The exact evaluation time at
which this architecture becomes reasonable depends on the amount of
communication overhead in the parallel implementation and the number of
particles in the swarm.  For our implementation we found that time to be around
100 milliseconds for swarms of 50 particles.  When the swarm size increases,
the minimum evaluation time at which this architecture should be used
decreases.

@@

Just as serial PSO papers do not report actual running times of their specific
implementations, we do not report running times, favoring the abstraction of
time steps.  However, given the time required for the evaluation of the
objective function and the communication overhead per iteration for a specific
implementation, a running time can be estimated from all of our results.
Simply multiply the number of time steps by the sum of the function evaluation
time and the overhead.  We wish to stress that the ``time steps'' we report are
proportional to wall-clock time, given the architecture we have assumed.

@@

We find benchmark functions useful in comparing algorithms, even though they
only take on the order of 1 millisecond to evaluate instead of the 100s that we
said were required to justify our parallel architecture.  The purpose of
exploring the behavior of an algorithm on benchmark functions, whether in
serial or in parallel optimization research, is because benchmark functions are
believed to represent functions that one might encounter in the real world,
regardless of the amount of time those functions take to evaluate.  While
benchmark functions themselves take fractions of a second to evaluate and thus
have no need of the kind of parallelization we explore in this paper, they also
have no need of particle swarm optimization at all, as they can be solved
analytically.  Yet they are useful to us and to optimization researchers
generally because they stand as surrogates for the kinds of functions
practitioners are actually interested in, and they allow us to explore the
behavior of optimization algorithms on several different classes of functions
in an easy and standardized way.

To provide additional evidence that performance on benchmark functions roughly
corresponds to performance on real-world problems, we also tested our
parallelization methods on a typical research problem, that of fitting a model
to a large quantity of data.  We generated 10,000 data points from a radial
basis function network with 10 bases and some added Gaussian noise.  We then
fit a radial basis function network to the data using PSO.  This amounted to a
30 dimensional function to optimize, with a function evaluation time on the
order of two seconds.  We refer to this function as ``the model fitting
problem.''

@@

\subsubsection{Sphere}
\label{sec:sphere20dims}

@@

Note here that \fig{griewank-manyiters} is interesting to compare to
\fig{basic-griewank1}.  The ordering of the methods in terms of the number of
time steps to completion is the same in both figures.  What is different is
that at 240 processors, most of our methods fail to find the optimum 100\% of
the time, while at 800 processors, all but Many Iterations succeed 100\% of the
time, and Many Iterations is very close.  At some point between 240 and 800
processors, SEPSO and Pick Best become successful, and at that point it is by
far better to use speculative methods than standard parallelizations.  At some
number above 800 processors, Many Iterations will be become successful and will
be the preferred method.

To further show this point, we present results using 1920 processors on the
20-dimensional Griewank function in \fig{griewank-all}.  With 1920 processors
Many Iterations finds the optimum 100\% of the time.  What is interesting is
that while going from 240 processors to 1920 processors, the standard method of
parallelization only decreased its time to completion by 5.9\%.  That is, the
extra 1680 processors in standard parallelizations provide \emph{no appreciable
benefit}.  In contrast, Many Iterations with 1920 processors has the same swarm
size of 240 particles and uses the additional processors for speculative
evaluation, giving a decrease in time to completion of 78.9\%, or a factor of
almost 5.  Once there are enough particles in the swarm to guarantee success
when optimizing a function, the best use of additional processors is to
speculate as far ahead as possible, speeding up the progress of the algorithm.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewankall}
  \caption{Function Griewank with 20 dimensions, comparing the performance of
  standard and speculative parallelizations as the number of processors
  increases.  All algorithms use the Ring topology.}
  \label{fig:griewank-all}
\end{figure}

@@

\subsection{Bare Bones PSO}
\label{sec:bare-results}

In order to demonstrate that our methods are applicable across PSO variants, we
ran experiments with Bare Bones PSO.  For this simple experiment we use the
Sphere function with 20 dimensions, taking the best Standard algorithm and the
best speculative algorithm on that function, which in this case is Standard
Complete and Pick Best Pruned Complete, as shown in \sec{sphere20dims}.
\fig{bare-bones} shows the results of running each of those algorithms with the
Bare Bones PSO equations.  We see a decrease in time to completion between
Standard and speculative parallelizations of 40\%.  While this is not quite as
large a speed up as the 60\% seen with constricted PSO, it is still a
significant decrease in time.  The key point of this experiment is that our
methods still show improvements over Standard parallelizations in PSO variants
other than the standard constricted PSO.  Our intent is not to give an
exhaustive listing of PSO variants and compare parallelization strategies with
each one; instead we offer evidence supporting our claim that speculative
evaluation should see similar increases in performance across most or all PSO
variants.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{bare_bones}
  \caption{Results for optimizing the 20-dimensional Sphere function with Bare
  Bones PSO.  We use 240 processors for each method, so each time step
  corresponds to 240 function evaluations.}
  \label{fig:bare-bones}
\end{figure}
